Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-14 Manuscript/sec2-review-of-graph-neural-network.tex
update section 2.

We follow the message-passing model \cite{gilmer_messgae_passing} to formally define a graph neuron.
The message-passing model is widely used in the cutting-edge GNN training systems like PyTorch Geometric (PyG) \cite{PyG} and Deep Graph Library (DGL) \cite{DGL}.
\figurename~\ref{fig:graph_neuron_structure} shows the structure of a graph neuron in the message-passing model.
Graph neurons at layer $l$ are made of three \emph{differentiable} functions: $\phi^l$, $\Sigma^l$ and $\gamma^l$.
The graph neuron calculates the output hidden vector $\boldsymbol{h}^{l+1}_i$ by
$$
\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}{\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j,	\boldsymbol{e}_{j,i})}).
$$

$\phi^l$ is the \emph{message} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ of the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ outputs a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For $v_i$, the message vectors $\boldsymbol{m}^l_{x,j}$ with $v_x \in \mathcal{N}(v_i)$ are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex update} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentialble.

@@

\begin{table}
	\hspace{-5em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^{(k)} = e_{j, i}\boldsymbol{z}_j^{(k-1)}$ &
		$O(d_{in})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		GraphSAGE-pool &
		Non-spectral &
		max &
		$\boldsymbol{m}_{j, i}^l =  \delta(\textcolor{blue}{\boldsymbol{W}^l_{pool}} \boldsymbol{h}_j^l + \textcolor{blue}{b}^l)$  &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{x}_j \parallel \boldsymbol{h}_j^l]$ &
		$O(f+d_{in})$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			& \text{Multi-head concatenation}: \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l}) \\
			& \text{Multi-head average}: \boldsymbol{m}_{j, i}^l = \frac{1}{K} \sum_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m, d_v) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer. Because two sum operator are interchangeable and sum operator and concatenation operator are interchangeable, we do it for GAT to explain more clearly in edge computations functions.}
	\label{tab:gnn_overview_edge}
\end{table}

@@

\textbf{GGNN} ()

>>>>

We follow the message-passing model \cite{gilmer_messgae_passing} to formally define a graph neuron.
The message-passing model is widely used in the cutting-edge GNN training systems like PyTorch Geometric (PyG) \cite{PyG} and Deep Graph Library (DGL) \cite{DGL}.
\figurename~\ref{fig:graph_neuron_structure} shows the structure of a graph neuron in the message-passing model.
A graph neuron collects messages from its neighbor graph neurons and outputs a hidden feature vector according to the aggregated messages.
Graph neurons at layer $l$ are made of three \emph{differentiable} functions: $\phi^l$, $\Sigma^l$ and $\gamma^l$.
The graph neuron calculates the output hidden vector $\boldsymbol{h}^{l+1}_i$ by
$$
\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}{\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j,	\boldsymbol{e}_{j,i})}).
$$

$\phi^l$ is the \emph{message} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ of the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ emits a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For $v_i$, the message vectors $\boldsymbol{m}^l_{x,j}$ of its adjacent edges ($v_x \in \mathcal{N}(v_i)$) are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex update} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentialble.

@@

\begin{table}
	\hspace{-5em}
    \begin{footnotesize}
        \begin{tabular}{ccp{8em}p{22em}r}
		\toprule
		GNN &
		Type &
		$\Sigma$ &
		$\phi$ &
		Complexity \\ \midrule
		ChebNet \cite{defferrad2016_chebnet} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^{(k)} = e_{j, i}\boldsymbol{z}_j^{(k-1)}$ &
		$O(d_{in})$ \\
		\textbf{GCN} \cite{kipf2017_gcn} &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = e_{j, i} \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		AGCN &
		Spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \tilde{e}_{j, i}^l \boldsymbol{h}_j^l$ &
		$O(d_{in})$ \\
		GraphSAGE &
		Non-spectral &
		mean/LSTM &
		$\boldsymbol{m}_{j, i}^l =  \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		GraphSAGE-pool &
		Non-spectral &
		max &
		$\boldsymbol{m}_{j, i}^l =  \delta(\textcolor{blue}{\boldsymbol{W}^l_{pool}} \boldsymbol{h}_j^l + \textcolor{blue}{b}^l)$  &
		$O(d_{in} * d_{out})$ \\
		Neural FPs &
		Non-spectral &
		sum &
		$\boldsymbol{m}_{j, i}^l = \boldsymbol{h}_j^l$ &
		$O(1)$ \\
		SSE &
		Recurrent &
		sum &
		$\boldsymbol{m}_{j, i}^l = [\boldsymbol{x}_j \parallel \boldsymbol{h}_j^l]$ &
		$O(f+d_{in})$ \\
		\textbf{GGNN} &
		Gated &
		sum &
		$\boldsymbol{m}_{j, i} = \textcolor{blue}{\boldsymbol{W}^l} \boldsymbol{h}_j^l$ &
		$O(d_{in} * d_{out})$ \\
		\textbf{GAT}&
		Attention &
		sum  &
		\begin{scriptsize}
			$\begin{aligned}[t]
			& \alpha_{j, i}^k = \frac {\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_i^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l] ))} {\sum_{k \in \mathcal{N}(i)}\exp(LeakyReLU(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}}  \boldsymbol{h}_k^l] ))} \\
			& \text{Multi-head concatenation}: \boldsymbol{m}_{j, i}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l}) \\
			& \text{Multi-head average}: \boldsymbol{m}_{j, i}^l = \frac{1}{K} \sum_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}} \boldsymbol{h}_j^{l})
			\end{aligned}$
		\end{scriptsize} &
		$O(K * d_{in} * d_{out})$ \\
		\textbf{GaAN} &
		Attention &
		sum,max,mean &
		\begin{scriptsize}
			$\begin{aligned}[t]
			&\alpha_{j, i}^k = \frac {\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l, k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}} \boldsymbol{h}_i^l] )} {\sum_{k \in \mathcal{N}(j)}\exp(\textcolor{blue}{\boldsymbol{a}^T} [ \textcolor{blue}{\boldsymbol{W}^{l,k}_{xa}} \boldsymbol{h}_j^l \parallel \textcolor{blue}{\boldsymbol{W}^{l,k}_{ya}}  \boldsymbol{h}_k^l] )} \\
			&\boldsymbol{m}_{j, i, 1}^l = \parallel_{k=1}^K \delta(\alpha_{j, i}^k \textcolor{blue}{\boldsymbol{W}^{l,k}_v} \boldsymbol{h}_j^{l}) \\
			&\boldsymbol{m}_{j, i, 2}^l = \textcolor{blue}{\boldsymbol{W}^l_m} \boldsymbol{h}_j^{l} \\
			&\boldsymbol{m}_{j, i, 3}^l = \boldsymbol{h}_j^l
			\end{aligned}$
		\end{scriptsize} &
		$O(max(d_a, d_m, d_v) * K * d_{in})$ \\
		\bottomrule
	\end{tabular}
\end{footnotesize}
	\caption{Typical graph neural networks and their edge computation functions.$d_{in}$ and $d_{out}$ are dimensions of the input and output hidden feature vectors, respectively.
		Blue variables are model parameters to learn. For ChebNet, we report its GNN sub-layer. Because two sum operator are interchangeable and sum operator and concatenation operator are interchangeable, we do it for GAT to explain more clearly in edge computations functions.}
	\label{tab:gnn_overview_edge}
\end{table}

@@

\textbf{GGNN} (high vertex \& low edge computational complexity):
GGNN introduces the gated recurrent unit (GRU) into the graph neural networks.
The vertex update function $\phi$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.

\textbf{GAT} (low vertex \& high edge computational complexity):
GAT introduces the attention mechanism into the graph neural networks.
During the message passing, it dynamically determines the weight for an edge based on the hidden feature vectors of its two end vertices.
Moreover, it adopts the multi-head mechanism that generates $K$ independent views for an edge, where $K$ is a hyperparameter.
In its message function $\phi$, $\$
